### 2015-02-27 src/4_OnlineLearning.tex
demos listas

@@

\begin{eqnarray*}
\mathbf{\phi}_*&=&(\mathbf{X}^\top \mathbf{X}+\lambda
\mathbb{I})^{-1}\mathbf{X}^\top y \\
\mathbf{\phi}_* &=& \displaystyle \big (\sum_{t=1}^m
\mathbf{x}_t \mathbf{x}_t  ^\top + \lambda
\mathbb{I}\big)^{-1} \sum_{t=1}^m y_t
\mathbf{x}_t \\ \\
\mathbf{\phi}_*&=& \mathbf{A}^{-1}\mathbf{b}
\end{eqnarray*}

\noindent where $\mathbf{A}=\mathbf{X}^\top \mathbf{X} +\lambda
\mathbb{I}$ and $\mathbf{b}=\displaystyle  \sum_{t=1}^m y_t \mathbf{x}_t$.

>>>>

\subsection{Ridge Regression}

One example of regularisation in linear models is Ridge Regression (RR), which is a regularised least squares method.
The least squares (LS) method is a well known way to solve a regression
problem. 
LS method consists of minimising the sum of squared errors:

\begin{eqnarray*}
\label{eq:lsm}
 J(\mathbf{X}) =& \displaystyle \sum_{t=1}^N (f(\mathbf{x}_t)-y_t)^2 \\
 =&\displaystyle \sum_{t=1}^N (\mathbf{\mathbf{X}}^\top {\mathbf{x}}_t-y_t)^2 \\
 =& \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{Y} \|_2^2 
\end{eqnarray*}

\noindent which is equivalent to find a solution to 
\begin{equation}
\label{eq:regproblem}
\underset{m \times n}{\mathbf{A}} \enskip \underset{n \times
l}{\mathbf{\mathbf{X}}} = \underset{m \times l}{\mathbf{Y}}
\end{equation}

The optimal solution for $\mathbf{\mathbf{X}}$ is:
\begin{equation*}
\mathbf{\mathbf{X}}= \mathbf{A}^+ \mathbf{Y}
%(\mathbf{X}^\top\mathbf{X}
%\mathbf{X})^{-1}\mathbf{X}^\top \mathbf{Y}
\end{equation*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\noindent where $\mathbf{A}^+$ is the Moore-Penrose pseudo-inverse
which can be written as follows:

\begin{equation}
\label{eq:pseudoinverse}
\mathbf{A}^+= (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \, .
\end{equation}

\textbf{Demo}\quad

To solve equation~(\ref{eq:pseudoinverse}) is equivalent to solve the following
optimisation problem:

\begin{equation}
\label{eq:RRproblem2}
\underset{\mathbf{X}(\lambda)}{\text{min}} \quad \|
\mathbf{C}\mathbf{\mathbf{X}} - \mathbf{F} \|_2^2
\end{equation}

\noindent where

\begin{equation*}
	\mathbf{C}=
	\begin{bmatrix} \quad &\mathbf{A}& \quad \\ \hdashline \quad
& \lambda\mathbb{I}&\quad  
\end{bmatrix} 
\quad \text{and} \quad
	\mathbf{F}=\begin{bmatrix} \quad &\mathbf{Y}& \quad \\ \hdashline \quad
& \mathbf{0}&\quad  \end{bmatrix} \, .
\end{equation*}

Applying equation~(\ref{eq:MP}) and considering that $\mathbf{C}^\top
\mathbf{C} = \mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I}$ and 
$\mathbf{C}^\top \mathbf{F}=\mathbf{A}^\top \mathbf{Y} $ we have:

\begin{eqnarray*}
\mathbf{X}(\lambda)&=&(\mathbf{C}^\top
\mathbf{C})^{-1}\mathbf{C}^\top \mathbf{F} \\
&=& (\mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I})^{-1} \mathbf{A}^\top \mathbf{Y}
\end{eqnarray*}

$\blacksquare$

However, when $\mathbf{A}$ is not full rank, i.e
$rank(\mathbf{A})=k <  n \leq m$, $\mathbf{A}^\top \mathbf{A}$ is
always singular and equation~(\ref{eq:pseudoinverse}) cannot be used.
More generally, the pseudo-inverse is best computed using the compact
singular value decomposition (SVD) of $\mathbf{A}$:

\begin{equation}
    \label{eq:compactsvd}
    \underset{m \times n}{\mathbf{A}}=
    \underset{m \times k}{\mathbf{U_1}} \enskip
    \underset{k \times k}{\mathbf{\Sigma_1}} \enskip
    \underset{k \times n}{\mathbf{V_1}^\top}
\end{equation}

\noindent as follows

\begin{equation}
\label{eq:pseudoinversesvd}
\mathbf{A}^+ = \mathbf{V_1\Sigma_1^{-1}U_1^\top}
\end{equation}

\textbf{Demo}\quad

In the case matrix $\mathbf{A}$ is non singular, its solution is
straight forward:

\begin{equation}
\label{OLSsolution}
    \mathbf{\mathbf{X}}=\mathbf{A}^{-1}\mathbf{Y}
\end{equation}

Since the problem shown in equation~(\ref{eq:regproblem}) has not
solution, the minimum norm given by equation~(\ref{OLSsolution}) is
obtained by solving the equivalent problem:

\begin{equation*}
\label{eq:proyectorsol}
\mathbf{X \hat{\mathbf{X}} = PY} 
\end{equation*}

\noindent where $\mathbf{P=U_1 U_1^\top}$ is the projection onto the
Col($\mathbf{A}$).

Since $\mathbf{V} = [\underset{(n \times k)}{\mathbf{V_1}} |
\underset{(n \times k)}{\mathbf{V_2}}]$ and $\mathbf{V_1^\top V_2 =
0}$ we can express $\mathbf{\hat{\mathbf{X}}} = \mathbf{V_1 \mathbf{X}_1 + V_2 \mathbf{X}_2}$
with $\mathbf{\mathbf{X}_2=0}$ because $\mathbf{\hat{\mathbf{X}}}$ lives in the
$\text{Row}(\mathbf{A})$ given by $\mathbf{V_1}$, so we have:

\begin{eqnarray*}
\mathbf{X \hat{\mathbf{X}}} &=& \mathbf{PB} \\
\mathbf{U_1 \Sigma_1 V_1^\top \hat{\mathbf{X}}} &=& \mathbf{U_1 U_1^\top B} \\
\mathbf{ V_1^\top \hat{\mathbf{X}}} &=&  \mathbf{\Sigma_1^{-1} U_1^\top B} \\ 
\mathbf{ V_1^\top V_1 \mathbf{X}_1} &=& \mathbf{\Sigma_1^{-1}
U_1^\top B} \\
\mathbf{\mathbf{X}_1}&=& \mathbf{\Sigma_1^{-1} U_1^\top B}
\end{eqnarray*}

\noindent from this result we can obtain $\mathbf{\hat{\mathbf{X}}}$ and
therefore the pseudo-inverse expression:

\begin{eqnarray*}
\mathbf{\hat{\mathbf{X}}} &=& \mathbf{V_1 \mathbf{X}_1} \\
                &=& \mathbf{V_1 \Sigma_1^{-1} U_1^\top Y} \\
\mathbf{X^+} &=& \mathbf{V_1 \Sigma_1^{-1} U_1^\top} \, .
\end{eqnarray*}

$\blacksquare$

In order to avoid the singularity of the matrix $\mathbf{A}^\top \mathbf{A}$, a regularisation term is introduced:

\begin{equation*}
\label{eq:problem} 
J(\mathbf{\mathbf{X}}) =  \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{Y} \|_2^2  + \lambda
 \| \mathbf{\mathbf{X}}\| ^2
\end{equation*}

\noindent which optimal solution $\mathbf{\mathbf{X}}_*$ is well known:

\begin{equation*}
\label{eq:optsolRR}
\mathbf{\mathbf{X}}_*=(\mathbf{A}^\top \mathbf{A}+\lambda \mathbb{I})^{-1}\mathbf{A}^\top y \, ,
\end{equation*}

\textbf{Demo}\quad

\begin{eqnarray*}
J(\mathbf{\mathbf{X}}) &=&  \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{Y} \|_2^2  + \lambda
 \| \mathbf{\mathbf{X}}\| ^2 \\
 &=&  \sum_{t=1}^N (\mathbf{\mathbf{X}}^\top {\bf a}_t-y_t)^2 + \lambda \sum_{i=1}^p \mathbf{X}_i^2 \\
 &=& (\mathbf{X}^\top a_1-y_1)^2 + \dots + (\mathbf{X}^\top a_N-y_N)^2 + \lambda (\mathbf{X}_1^2+\dots+\mathbf{X}_p^2)
\end{eqnarray*}
\noindent taking derivatives
 \begin{eqnarray*}
 \frac{\partial J(\mathbf{\mathbf{X}})}{\partial \mathbf{X}_1}&=& 
 2(\mathbf{X}^\top \mathbf{a}_1-y_1)\mathbf{a}_{11} + \dots + 2(\mathbf{X}^\top \mathbf{a}_N-y_N)\mathbf{a}_{N1} + 2\lambda \mathbf{X}_1 \\
 &=& 2\mathbf{a}_1^\top(\mathbf{A}\mathbf{X}-\mathbf{Y}) + 2\lambda\mathbf{X}_1\\
& \vdots &\\
  \frac{\partial J(\mathbf{\mathbf{X}})}{\partial \mathbf{X}_p}&=& 
 2\mathbf{a}_p^\top(\mathbf{A}\mathbf{X}-\mathbf{Y}) + 2\lambda\mathbf{X}_p 
\end{eqnarray*}

Then we have that:
\begin{eqnarray*}
\frac{\partial J(\mathbf{\mathbf{X}})}{\partial \mathbf{X}}&=& 
 2\mathbf{A}^\top(\mathbf{A}\mathbf{X}-\mathbf{Y}) + 2\lambda\mathbf{X}
\end{eqnarray*}
Since $\frac{\partial J(\mathbf{\mathbf{X}})}{\partial \mathbf{X}}=0$ we have:
\begin{eqnarray*}
2\mathbf{A}^\top(\mathbf{A}\mathbf{X}-\mathbf{Y}) + 2\lambda\mathbf{X}&=&0 \\
\mathbf{A}^\top\mathbf{A}\mathbf{X} - \mathbf{A}^\top\mathbf{Y} + \lambda\mathbf{X} &=& 0\\
(\mathbf{A}^\top\mathbf{A}+\lambda\mathbb{I})\mathbf{X} &=&  \mathbf{A}^\top\mathbf{Y} \\
\mathbf{X} &=& (\mathbf{A}^\top\mathbf{A}+\lambda\mathbb{I})^{-1}  \mathbf{A}^\top\mathbf{Y}
\end{eqnarray*}

$\blacksquare$

When $\mathbf{A}$ is orthonormal then $\mathbf{A}^\top \mathbf{A} =\mathbb{I}$ and there is a simple relation between the ridge estimator and the OLS estimator:
\begin{eqnarray*}
\mathbf{X}_* (\lambda) &=& (\mathbf{A}^\top \mathbf{A}+\lambda \mathbb{I})^{-1}\mathbf{A}^\top \mathbf{Y} \\
 &=& (\mathbb{I} + \lambda \mathbb{I})^{-1} \mathbf{A}^\top \mathbf{Y} \\
 &=&(1+\lambda)^{-1} \mathbb{I} \mathbf{A}^\top \mathbf{Y} \\
 &=&(1+\lambda)^{-1} (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{Y} \\
 &=&(1+\lambda)^{-1} \mathbf{X}
\end{eqnarray*}

@@

\begin{eqnarray*}
\mathbf{\mathbf{X}}_*&=&(\mathbf{A}^\top \mathbf{A}+\lambda
\mathbb{I})^{-1}\mathbf{A}^\top y \\
\mathbf{\mathbf{X}}_* &=& \displaystyle \big (\sum_{t=1}^m
\mathbf{x}_t \mathbf{x}_t  ^\top + \lambda
\mathbb{I}\big)^{-1} \sum_{t=1}^m y_t
\mathbf{x}_t \\ \\
\mathbf{\mathbf{X}}_*&=& \mathbf{A}^{-1}\mathbf{b}
\end{eqnarray*}

\noindent where $\mathbf{A}=\mathbf{A}^\top \mathbf{A} +\lambda
\mathbb{I}$ and $\mathbf{b}=\displaystyle  \sum_{t=1}^m y_t \mathbf{x}_t$.
